Manuscript title here

### 2023-09-04 manuscript.Rmd
Update manuscript.Rmd

Injury is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].

@@

Study design
------------

<!-- Present key elements of study design -->

>>>>

Injury is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021]. To combat this, dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

Trauma teams are multidisciplinary and operate in these centers. They consist of individuals from the specialities of emergency medicine, surgery, intensive care, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency physicians are often the personnel activating the trauma code using anatomical and physiological criterias, mechanism of injury and information obtained from emergency service personnel.

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. In the case of a delayed trauma team activation (DTTA) it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994]. Studies have also shown that different factors such as age over 55, nonwhite ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].

This study will assess delayed treatment at the Karolinska University Hospital in Solna between the years 2012 and 2022 through the trauma registry and trauma care quality databse. Our objectives are to analyse how different factors are associated with delayed treatment in patients.

@@

Study design
------------
This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.
